Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-21 linear_adia.tex
regular update, added analytical calc of critical timescale for adia vsi

\subsubsection{Qualitative condition for buoyancy to be negligible in
  the presence of thermal relaxation}
{\bf this section might be wrong, but keep for reference for now}
We have shown that when $\gamma>1$ so that there is a positive
vertical entropy gradient and perturbations evolve
adiabatically ($t_c\to\infty$) that the VSI is strongly
surpressed. This is the limit where the term proportional to
$t_c^{-1}$ in the linearized energy equation (Eq. \ref{lin_energy})
becomes negligible compared to the term proportional to
$\gamma/\Gamma-1$. The latter term is associated with the vertical
entropy gradient, or buoyancy frequency squared, $N_z^2$.

Reversing the argument, we expect to recover the VSI when the bouyancy
frequency becomes negligible compared to the cooling rate, or 
\begin{align}
  t_c \ll \frac{\Omega}{\gamma N_z^2}. 
\end{align} 
Parametrizing $t_c = \beta\Omega_k^{-1}$ and inserting the expression
for $N_z^2$ appropriate for vertically isothermal disks, we expect
\begin{align}
  \beta \ll
  \frac{1}{H^2_\mathrm{iso}\left(\gamma-1\right)\left(\frac{\p\ln\rho}{\p
          z}\right)^2} 
\end{align}
is required to enable the VSI, where we used $\Omega\sim\Omega_k$.  
This inequality is satisfied at the
midplane and as $|z|\to\infty$. It can be satisifed at all heights if
it is satisfied at the height of $\mathrm{max}|\p\ln\rho/\p z|$. 
Using the full expression for the density field for vertically isothermal
disks (Eq. \ref{vertiso_eqm}), we expect 
\begin{align}
  \beta \ll \frac{27}{4(\gamma-1)}\epsilon^2,
\end{align}
i.e. $\beta\ll O(\epsilon^2)$, is a sufficient condition for
buoyancy to be negligible.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/rate_theory_grow_relax}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/rate_theory_grow_relax2}
  \caption{Growth rate of the fundamental VSI mode in a disk with
    dimensionless thermal relaxation timescales
    $\beta\in[10^{-3},1]$ (top) and $\beta\in[1,10^3]$ (bottom).
Other disk disk parameters are
    $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. The growth rates are
    calculated from the dispersion relation Eq. \ref{relax_disp}. 
    \label{relax_disp_fig}}  
\end{figure}

@@

\subsubsection{Upper limit on the thermal relaxation timescale for the
  fundamental isothermal VSI mode}
Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
fundamental isothermal VSI tends to a constant for large
$\khat$. Using this clue, we can calculate this critical $\beta$ by
considering $\khat\to\infty$. We proceed as follows.

@@

>>>>

% \subsubsection{Qualitative condition for buoyancy to be negligible in
%   the presence of thermal relaxation}
% {\bf this section might be wrong, but keep for reference for now}
% We have shown that when $\gamma>1$ so that there is a positive
% vertical entropy gradient and perturbations evolve
% adiabatically ($t_c\to\infty$) that the VSI is strongly
% surpressed. This is the limit where the term proportional to
% $t_c^{-1}$ in the linearized energy equation (Eq. \ref{lin_energy})
% becomes negligible compared to the term proportional to
% $\gamma/\Gamma-1$. The latter term is associated with the vertical
% entropy gradient, or buoyancy frequency squared, $N_z^2$.

% Reversing the argument, we expect to recover the VSI when the bouyancy
% frequency becomes negligible compared to the cooling rate, or 
% \begin{align}
%   t_c \ll \frac{\Omega}{\gamma N_z^2}. 
% \end{align} 
% Parametrizing $t_c = \beta\Omega_k^{-1}$ and inserting the expression
% for $N_z^2$ appropriate for vertically isothermal disks, we expect
% \begin{align}
%   \beta \ll
%   \frac{1}{H^2_\mathrm{iso}\left(\gamma-1\right)\left(\frac{\p\ln\rho}{\p
%           z}\right)^2} 
% \end{align}
% is required to enable the VSI, where we used $\Omega\sim\Omega_k$.  
% This inequality is satisfied at the
% midplane and as $|z|\to\infty$. It can be satisifed at all heights if
% it is satisfied at the height of $\mathrm{max}|\p\ln\rho/\p z|$. 
% Using the full expression for the density field for vertically isothermal
% disks (Eq. \ref{vertiso_eqm}), we expect 
% \begin{align}
%   \beta \ll \frac{27}{4(\gamma-1)}\epsilon^2,
% \end{align}
% i.e. $\beta\ll O(\epsilon^2)$, is a sufficient condition for
% buoyancy to be negligible.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/rate_theory_grow_relax}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/rate_theory_grow_relax2}
  \caption{Growth rate of the fundamental VSI mode in a disk with
    dimensionless thermal relaxation timescales
    $\beta\in[10^{-3},1]$ (top) and $\beta\in[1,10^3]$ (bottom).
Other disk parameters are
    $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. The growth rates are
    calculated from the dispersion relation Eq. \ref{relax_disp}. 
    \label{relax_disp_fig}}  
\end{figure}

@@

\subsubsection{Upper limit on the thermal relaxation timescale for the
  fundamental isothermal VSI mode}\label{iso_vsi_beta_crit}
Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
fundamental isothermal VSI tends to a constant for large
$\khat$. Using this clue, we can calculate this critical $\beta$ by
considering $\khat\to\infty$. We proceed as follows.

@@

\subsubsection{Lower limit on the thermal relaxation timescale for the
  fundamental adiabatic VSI mode}
{\bf Note: for reference only. Not to be included in paper.} 
Fig. \ref{relax_bcrit} indicates there exists a lower limit on the
thermal relaxation time for the system to behave adiabatically and
support the fundamental adiabatic VSI. This critical
timescale satisfies Eq. \ref{relax_cond1}---\ref{relax_cond2} with 
\begin{align}
  \beta \sim \beta_0\khat^{1/2},\quad \hat{\omega}  \sim
  \hat{\omega}_0\khat^{-1/2} \text{ as } \khat\to\infty.
\end{align}
The constants $\beta_0$ and $\hat{\omega}_0$ are determined by the
simultaneous equations
\begin{align}
  & 0 = \gamma\beta_0^2(1-\gamma) + 2\epsilon q \gamma
  \beta_0\hat{\omega}_0 - \hat{\omega}_0^2 +
  \beta_0^2\gamma^2\hat{\omega}_0^4.\\
  & 0= \beta_0(1-\gamma) + \epsilon q \hat{\omega}_0 - \epsilon q
  \gamma^2 \beta_0^2\hat{\omega}_0^3. 
\end{align}
For $\epsilon = 0.05$, $q=-1$ and $\gamma=1.4$, the appropriate
solution is $\beta_0 = 10.33$ and $\hat{\omega}_0 = 0.74$.

### 2015-01-21 numerical.tex
restructing analytic discussion
regular update, added analytical calc of critical timescale for adia vsi

Unless otherwise stated, our fiducial disk model is a
nearly-vertically isothermal disk with $\Gamma=1.011$, $p=-1.5$,
$q=-1$, $\epsilon=0.1$, and the perturbation wavenumber $k_x = 200\pi/r$
(or $\hat{k}=20\pi$).

@@

We have also solved this problem with in the low-frequency 
approximation by setting $D\to\kappa^2$ in Eq. \ref{iso_ode}. We
obtained very similar results, as expected for modes with
$|\sigma/\Omega_k|\ll1$.

@@

\subsubsection{Other adiabatic indices}
In Fig. \ref{adia_growth_num} we plot the eigenfrequencies of the
fundamental VSI mode for a range of $\gamma$ values. This plot is to
be compared with Fig. \ref{adia_growth}.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/compare_eigen_real1}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/compare_eigen_imag1}
  \caption{Real frequency (top) and growth rate 
    (bottom) of the fundamental VSI mode in nearly-vertically   
    isothermal disks ($\Gamma=1.011$) with $\epsilon=0.1$ and  
    $q=-1$, subject to adiabatic perturbations
    ($\beta=10^9$). The eigenfrequencies are calculated by numerically
    solving the full linear eigenvalue problem. \label{adia_growth_num}}   
\end{figure}

\subsection{Effect of thermal relaxation}
We now introduce thermal relaxation. Here we consider a slightly
different disk model with $\epsilon=0.05$ and
$\gamma=1.4$. This model was used by \citep{nelson13}, where the
authors find a cooling timescale of $\beta\geq 0.1$ was sufficient to
stabilize the VSI in non-linear hydrodynamic simulations.

% \subsection{Vertically non-isothermal disks}
% {\bf big gamma larger than 1 but equal to small gamma, with tcool}

% \subsection{Non-uniform thermal relaxation} 
% {\bf variable tcool, large at midplane, small in atmosphere. simplest
%   model is to multiply tcool by background density}

>>>>

Unless otherwise stated, our fiducial disk model is a
nearly-vertically isothermal disk with $\Gamma=1.011$, $p=-1.5$,
$q=-1$, $\epsilon=0.1$, $\zmax = 10\epsilon r$, and the perturbation
wavenumber $k_x = 200\pi/r$ (or $\hat{k}=20\pi$).

@@

We have also solved this problem with in the low-frequency 
approximation by setting $D\to\kappa^2$ in Eq. \ref{ode_w} (with $Q=W$). We
obtained very similar results, as expected for modes with
$|\sigma/\Omega_k|\ll1$.

@@

\subsection{Other adiabatic indices}
In Fig. \ref{adia_growth_num} we plot the eigenfrequencies of the
fundamental VSI mode for $\gamma\in[1,2]$. To select the appropriate
eigenvalues we vary the parameters slowly away from a reference solution.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/compare_eigen_real1}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/compare_eigen_imag1}
  \caption{Real frequency (top) and growth rate 
    (bottom) of the fundamental VSI mode in nearly-vertically   
    isothermal disks ($\Gamma=1.011$) with $\epsilon=0.1$ and  
    $q=-1$, subject to adiabatic perturbations
    ($\beta=10^9$). The eigenfrequencies are calculated by numerically
    solving the full linear eigenvalue problem. This plot is to
    be compared with Fig. \ref{adia_growth}. \label{adia_growth_num}}   
\end{figure}

\subsection{Effect of thermal relaxation}
We now introduce thermal relaxation. Here we consider a slightly
different disk model with $\epsilon=0.05$ and $\gamma=1.4$, as used by
\cite{nelson13}. Fig. \ref{relax_growth_num} plots the growth rates
of the fundamental mode function of $k_x$. %corrugation mode
% We first obtain the eigenvalues for the
% fundamental mode for $k_xH_\mathrm{iso}=0$, then vary parameters
% slowly to obtain eigenvalues for other values of $k_x$ or $\beta$.

As above, Fig. \ref{relax_growth_num} compares well with
Fig. \ref{relax_disp_fig} except at $k_xH_\mathrm{iso}\lesssim 1$
where the low-frequency approximation fails. The non-smoothness in the
curve for $\beta=0.063$ at $k_xH_\mathrm{iso}\gtrsim50$ was found to be
associated with perturbations near the vertical boundaries. This is
likely attributable to the free surface boundary condition 
(cf. regularity condition at infinity in \S\ref{disp_relax}).

Fig. \ref{relax_growth_num} confirms our analytical discussion that
introducing a small but finite cooling time stabilizes the fundamental
VSI, but there is a preferable wavenumber around $k_xH_\mathrm{iso} =
O(10)$ that maximizes this effect.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/compare_eigen_imag_bloop2}
  \caption{Growth rates of the fundamental VSI mode in
    nearly-vertically isothermal disks ($\Gamma=1.011$) with
    $\epsilon=0.05$ and $q=-1$, with thermal relaxation timescale
    $\beta$. The eigenfrequencies are calculated by numerically
    solving the full linear eigenvalue problem. This plot is to
    be compared with Fig. \ref{relax_disp_fig}. \label{relax_growth_num}}   
\end{figure}

We visiualize the effect of thermal relaxation in
Fig. \ref{relax_eigenW_num} by comparing the eigenfunction $W$ for
$\beta=0.01$ and $\beta=0.125$, the latter is close to the critical
value beyond which the isothermal fundamental VSI is stabilized
(\S\ref{iso_vsi_beta_crit}).

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/eigenvectorW_beta0d01}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 0cm]{figures/eigenvectorW_beta0d125}
  \caption{Numerically-calculated eigenfunction of the fundamental VSI
    mode in a disk with dimensionless thermal relaxation timescale
    $\beta \equiv t_c\Omega_k = 0.01$ (top) and $0.125$ (bottom) for
    perturbations with a radial wavenumber $k_xH_\mathrm{iso}=10$. The
    real (imaginary) part of $W$ is plotted as the solid (dotted)
    line. Other disk parameters are
    $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$.
    \label{relax_eigenW_num}}  
\end{figure}

\citeauthor{nelson13} find a thermal relaxation timescale of
$\beta\gtrsim 0.6$ stabilized the VSI for this disk model in non-linear
hydrodynamic simulations. This is consistent with our numerical
results.
